---
title: "R Notebook"
output:
  html_document: default
  html_notebook: default
---

# Random Forests

We explore using random forests to identify which genes are most important in differentiating between eye tissue and other tissues. 

```{r}
# Variable to determine whether to use:
# FALSE: previously computed datasets
# TRUE: compute new datasets
freshRun = FALSE
```

We load in the necessary packages and data

```{r, message = FALSE}
library(dplyr)
library(ggplot2)
library(randomForest)
library(caret)
library(Rtsne)

load("~/Documents/git/unified_gene_expression/data/counts_processed.Rdata")
load("~/Documents/git/unified_gene_expression/data/core_metaData.Rdata")

expr = t(counts)
meta = core_tight

dim(expr)
dim(meta)
```

We left_join the meta sheet (for the *sample_accession* variable) and the expression matrix. We then subset the resulting matrix into *RPE*, *Retina*, and *RPE+Retina* tables.

```{r, message = FALSE}
if(freshRun){
  sample_accession = rownames(expr)
  expr.temp = as.data.frame(cbind(sample_accession, expr))
  
  rownames(expr.temp)[1:5]
  colnames(expr.temp)[1:5]
  
  expr.meta = left_join(expr.temp, meta, by = "sample_accession") %>% dplyr::select(-run_accession) %>% distinct()
  
  expr.temp$sample_accession = as.character(expr.temp$sample_accession)
  
  expr.temp$sample_accession[1:5]
  expr.meta$sample_accession[1:5]
  identical(expr.temp$sample_accession, expr.meta$sample_accession)
  
  rownames(expr.meta) = expr.meta$sample_accession
  
  expr.meta[1:5, 1:3]
  
  class(expr.meta)
  table(expr.meta$Tissue)
  
  expr.meta <- filter(expr.meta, Origin == "Tissue")
  
  class(expr.meta)
  table(expr.meta$Tissue)
  
  expr.retina = filter(expr.meta, Tissue == "Retina")
  expr.rpe = filter(expr.meta, Tissue == "RPE")
  expr.retina.rpe = filter(expr.meta, Tissue == "Retina" | Tissue == "RPE")
  
  expr.meta <- filter(expr.meta, Tissue != "Brain" & Tissue != "Cornea" & Tissue != "Retina" & Tissue != "RPE")
  
  set.seed(47)
  expr.comp.retina <- expr.meta %>% group_by(Tissue) %>% sample_n(size = 4) %>% ungroup() %>% as.data.frame()
  set.seed(47)
  expr.comp.rpe <- expr.meta %>% group_by(Tissue) %>% sample_n(size = 2) %>% ungroup() %>% as.data.frame()
  set.seed(47)
  expr.comp.retina.rpe <- expr.meta %>% group_by(Tissue) %>% sample_n(size = 6) %>% ungroup() %>% as.data.frame()
  
  # Eye tissues
  rownames(expr.retina) = expr.retina$sample_accession
  rownames(expr.rpe) = expr.rpe$sample_accession
  rownames(expr.retina.rpe) = expr.retina.rpe$sample_accession
  
  expr.retina = expr.retina[, c(2:ncol(expr.retina), 1)]
  expr.rpe = expr.rpe[, c(2:ncol(expr.rpe), 1)]
  expr.retina.rpe = expr.retina.rpe[, c(2:ncol(expr.retina.rpe), 1)]
  ###
  # Other tissues
  rownames(expr.comp.retina) = expr.comp.retina$sample_accession
  rownames(expr.comp.rpe) = expr.comp.rpe$sample_accession
  rownames(expr.comp.retina.rpe) = expr.comp.retina.rpe$sample_accession
  
  expr.comp.retina = expr.comp.retina[, c(2:ncol(expr.comp.retina), 1)]
  expr.comp.rpe = expr.comp.rpe[, c(2:ncol(expr.comp.rpe), 1)]
  expr.comp.retina.rpe = expr.comp.retina.rpe[, c(2:ncol(expr.comp.retina.rpe), 1)]
  ###
  
  dim(expr.retina)
  dim(expr.comp.retina)
  dim(expr.rpe)
  dim(expr.comp.rpe)
  dim(expr.retina.rpe)
  dim(expr.comp.retina.rpe)
  
  save(expr.retina, file = "~/Documents/git/unified_gene_expression/data/RF_comparisons/counts_retina.Rdata")
  save(expr.rpe, file = "~/Documents/git/unified_gene_expression/data/RF_comparisons/counts_rpe.Rdata")
  save(expr.retina.rpe, file = "~/Documents/git/unified_gene_expression/data/RF_comparisons/counts_retina_rpe.Rdata")
  save(expr.comp.retina, file = "~/Documents/git/unified_gene_expression/data/RF_comparisons/counts_composite_retina.Rdata")
  save(expr.comp.rpe, file = "~/Documents/git/unified_gene_expression/data/RF_comparisons/counts_composite_rpe.Rdata")
  save(expr.comp.retina.rpe, file = "~/Documents/git/unified_gene_expression/data/RF_comparisons/counts_composite_retina_rpe.Rdata")
}else{
  load("~/Documents/git/unified_gene_expression/data/RF_comparisons/counts_retina.Rdata")
  load("~/Documents/git/unified_gene_expression/data/RF_comparisons/counts_rpe.Rdata")
  load("~/Documents/git/unified_gene_expression/data/RF_comparisons/counts_retina_rpe.Rdata")
  load("~/Documents/git/unified_gene_expression/data/RF_comparisons/counts_composite_retina.Rdata")
  load("~/Documents/git/unified_gene_expression/data/RF_comparisons/counts_composite_rpe.Rdata")
  load("~/Documents/git/unified_gene_expression/data/RF_comparisons/counts_composite_retina_rpe.Rdata")
}

```

# Retina

We now build the random forest for the retina samples. We find that with $nTree = 1000$, the random forest perfectly classifies the samples. We find the 50 most important genes based on *MeanDecreaseAccuracy* and *MeanDecreaseGini*, and we find that 41 of the genes in these groups are the same.

```{r, message = FALSE}
#####################################
##         Retina RF               ##
#####################################

meta.retina.rf = expr.retina[, which(colnames(expr.retina) == "study_accession"):ncol(expr.retina)]
expr.retina.rf = expr.retina[, 1:(which(colnames(expr.retina) == "study_accession") - 1)]

meta.comp.rf = expr.comp.retina[, which(colnames(expr.comp.retina) == "study_accession"):ncol(expr.comp.retina)]
expr.comp.rf = expr.comp.retina[, 1:(which(colnames(expr.comp.retina) == "study_accession") - 1)]

expr.retina.rf[] = sapply(expr.retina.rf, as.numeric)
expr.comp.rf[] = sapply(expr.comp.rf, as.numeric)

meta.comp.rf$Tissue = rep("Composite", length(meta.comp.rf$Tissue))

# Combining data
meta.rf = bind_rows(meta.retina.rf, meta.comp.rf)
expr.rf = bind_rows(expr.retina.rf, expr.comp.rf)

# Random forest
set.seed(47)
rf.retina.comp <- randomForest(x = expr.rf, y = factor(meta.rf$Tissue), ntree = 1000, importance = TRUE, scale = FALSE)
rf.retina.comp

varImpPlot(rf.retina.comp, scale = TRUE, n.var = 50, cex = 0.45, pt.cex = 0.75, main = "Top 50 Most Important Genes")

impt.retina = importance(rf.retina.comp, scale = TRUE)

sum(impt.retina[, "MeanDecreaseAccuracy"] >= qnorm(0.95, mean = 0, sd = 1))

genes.acc = impt.retina[order(impt.retina[, "MeanDecreaseAccuracy"], decreasing = TRUE), ][1:50, ]
genes.gini = impt.retina[order(impt.retina[, "MeanDecreaseGini"], decreasing = TRUE), ][1:50, ]

length(intersect(rownames(genes.acc), rownames(genes.gini)))

sig.genes.list.retina = rownames(impt.retina)[which(impt.retina[, "MeanDecreaseAccuracy"] >= qnorm(0.95, mean = 0, sd = 1))]

write.table(sig.genes.list.retina, file = "~/Documents/git/unified_gene_expression/results/RF_MDA_sig_retina.txt", 
            sep = "\n", row.names = F, col.names = F)

```

# RPE

We now build the random forest for the RPE samples. We find that with $nTree = 1000$, the random forest only misclassified 3 samples. We find the 50 most important genes based on *MeanDecreaseAccuracy* and *MeanDecreaseGini*, and we find that all 50 of the genes in these groups are the same.

```{r, message = FALSE}
#####################################
##         RPE RF                  ##
#####################################

meta.rpe.rf = expr.rpe[, which(colnames(expr.rpe) == "study_accession"):ncol(expr.rpe)]
expr.rpe.rf = expr.rpe[, 1:(which(colnames(expr.rpe) == "study_accession") - 1)]

meta.comp.rf = expr.comp.rpe[, which(colnames(expr.comp.rpe) == "study_accession"):ncol(expr.comp.rpe)]
expr.comp.rf = expr.comp.rpe[, 1:(which(colnames(expr.comp.rpe) == "study_accession") - 1)]

expr.rpe.rf[] = sapply(expr.rpe.rf, as.numeric)
expr.comp.rf[] = sapply(expr.comp.rf, as.numeric)

meta.comp.rf$Tissue = rep("Composite", length(meta.comp.rf$Tissue))

# Combining data
meta.rf = bind_rows(meta.rpe.rf, meta.comp.rf)
expr.rf = bind_rows(expr.rpe.rf, expr.comp.rf)

# Random forest
set.seed(47)
rf.rpe.comp <- randomForest(x = expr.rf, y = factor(meta.rf$Tissue), ntree = 1000, importance = TRUE)
rf.rpe.comp

varImpPlot(rf.rpe.comp, scale = TRUE, n.var = 50, cex = 0.45, pt.cex = 0.75, main = "Top 50 Most Important Genes")

impt.rpe = importance(rf.rpe.comp, scale = TRUE)

sum(impt.rpe[, "MeanDecreaseAccuracy"] >= qnorm(0.95, mean = 0, sd = 1))

genes.acc = impt.rpe[order(impt.rpe[, "MeanDecreaseAccuracy"], decreasing = TRUE), ][1:100, ]
genes.gini = impt.rpe[order(impt.rpe[, "MeanDecreaseGini"], decreasing = TRUE), ][1:50, ]

length(intersect(rownames(genes.acc), rownames(genes.gini)))

sig.genes.list.rpe = rownames(impt.rpe)[which(impt.rpe[, "MeanDecreaseAccuracy"] >= qnorm(0.95, mean = 0, sd = 1))]
```

# Retina+RPE

We now build the random forest for Retina and RPE together. We find that with $nTree = 1000$, the random forest only misclassified 3 samples. We find the 50 most important genes based on *MeanDecreaseAccuracy* and *MeanDecreaseGini*, and we find that 41 of the genes in these groups are the same.

```{r, message = FALSE}
#####################################
##         Retina+RPE RF           ##
#####################################

meta.retina.rpe.rf = expr.retina.rpe[, which(colnames(expr.retina.rpe) == "study_accession"):ncol(expr.retina.rpe)]
expr.retina.rpe.rf = expr.retina.rpe[, 1:(which(colnames(expr.retina.rpe) == "study_accession") - 1)]

meta.comp.rf = expr.comp.retina.rpe[, which(colnames(expr.comp.retina.rpe) == "study_accession"):ncol(expr.comp.retina.rpe)]
expr.comp.rf = expr.comp.retina.rpe[, 1:(which(colnames(expr.comp.retina.rpe) == "study_accession") - 1)]

expr.retina.rpe.rf[] = sapply(expr.retina.rpe.rf, as.numeric)
expr.comp.rf[] = sapply(expr.comp.rf, as.numeric)

meta.comp.rf$Tissue = rep("Composite", length(meta.comp.rf$Tissue))
meta.retina.rpe.rf$Tissue = rep("Retina.RPE", length(meta.retina.rpe.rf$Tissue))

# Combining data
meta.rf = bind_rows(meta.retina.rpe.rf, meta.comp.rf)
expr.rf = bind_rows(expr.retina.rpe.rf, expr.comp.rf)

# Random forest
set.seed(47)
rf.retina.rpe.comp <- randomForest(x = expr.rf, y = factor(meta.rf$Tissue), ntree = 1000, importance = TRUE)
rf.retina.rpe.comp

varImpPlot(rf.retina.rpe.comp, scale = TRUE, n.var = 50, cex = 0.45, pt.cex = 0.75, main = "Top 50 Most Important Genes")

impt.retina.rpe = importance(rf.retina.rpe.comp, scale = TRUE)

sum(impt.retina.rpe[, "MeanDecreaseAccuracy"] >= qnorm(0.95, mean = 0, sd = 1))

genes.acc = impt.retina.rpe[order(impt.retina.rpe[, "MeanDecreaseAccuracy"], decreasing = TRUE), ][1:50, ]
genes.gini = impt.retina.rpe[order(impt.retina.rpe[, "MeanDecreaseGini"], decreasing = TRUE), ][1:50, ]

length(intersect(rownames(genes.acc), rownames(genes.gini)))

sig.genes.list.retina.rpe = rownames(impt.retina.rpe)[which(impt.retina.rpe[, "MeanDecreaseAccuracy"] >= qnorm(0.95, mean = 0, sd = 1))]
```

# Validating Random Forests

In the Leo Breiman implementation of random forests (the *randomForest* package we're using), they propose using the mean decrease in accuracy as a sort of permutation test statistic that you can map to a standard normal distribution. As such, for validating the genes that appear "most important" in the random forests, let's consider the genes for which their mean decrease in accuracy is statistically significant by permutation test. 

```{r, message = FALSE}
load("~/Documents/git/unified_gene_expression/data/lengthScaledTPM_processed.Rdata")
load("~/Documents/git/unified_gene_expression/data/core_metaData.Rdata")

core_info = core_tight

lengthScaledTPM <- lengthScaledTPM_qsmooth_highExp_remove_lowGenes

lengthScaledTPM <- lengthScaledTPM[,!(is.na(lengthScaledTPM[1,]))]

perplexities = seq(10, 50, by = 10)
############################
# Retina
############################
for(p in perplexities){
  # All genes
  set.seed(47)
  tsne_out <- Rtsne(as.matrix(log2(t(lengthScaledTPM)+1)),perplexity = p, check_duplicates = FALSE, theta=0.0 )

  tsne_plot <- data.frame(tsne_out$Y)
  tsne_plot$sample_accession <- colnames(lengthScaledTPM)

  p1 = tsne_plot %>% left_join(.,core_info)  %>%
    ggplot(.,aes(x=X1,y=X2,colour=Tissue,shape=Tissue)) + 
    geom_point(size=4) + scale_shape_manual(values=c(0:20,35:50)) +
    ggtitle(paste0("t-sne. Perplexity = ", p)) + 
    coord_cartesian(xlim = c(-50, 50), ylim = c(-50, 50))
  print(p1)
  
  # RF genes removed
  sig.genes.indices = which(rownames(lengthScaledTPM) %in% sig.genes.list.retina)
  lengthScaledTPM.sig.genes = lengthScaledTPM[-(sig.genes.indices), ]
  
  set.seed(47)
  tsne_out <- Rtsne(as.matrix(log2(t(lengthScaledTPM.sig.genes)+1)),perplexity = p, check_duplicates = FALSE, theta=0.0 )

  tsne_plot <- data.frame(tsne_out$Y)
  tsne_plot$sample_accession <- colnames(lengthScaledTPM.sig.genes)
  
  p2 = tsne_plot %>% left_join(.,core_info)  %>%
    ggplot(.,aes(x=X1,y=X2,colour=Tissue,shape=Tissue)) + 
    geom_point(size=4) + scale_shape_manual(values=c(0:20,35:50)) +
    ggtitle(paste0("t-sne. Perplexity = ", p)) + 
    coord_cartesian(xlim = c(-50, 50), ylim = c(-50, 50))
  print(p2)
}

############################
# RPE
############################
for(p in perplexities){
  # All genes
  set.seed(47)
  tsne_out <- Rtsne(as.matrix(log2(t(lengthScaledTPM)+1)),perplexity = p, check_duplicates = FALSE, theta=0.0 )

  tsne_plot <- data.frame(tsne_out$Y)
  tsne_plot$sample_accession <- colnames(lengthScaledTPM)

  p1 = tsne_plot %>% left_join(.,core_info)  %>%
    ggplot(.,aes(x=X1,y=X2,colour=Tissue,shape=Tissue)) + 
    geom_point(size=4) + scale_shape_manual(values=c(0:20,35:50)) +
    ggtitle(paste0("t-sne. Perplexity = ", p)) + 
    coord_cartesian(xlim = c(-50, 50), ylim = c(-50, 50))
  print(p1)
  
  # RF genes removed
  sig.genes.indices = which(rownames(lengthScaledTPM) %in% sig.genes.list.rpe)
  lengthScaledTPM.sig.genes = lengthScaledTPM[-(sig.genes.indices), ]
  
  set.seed(47)
  tsne_out <- Rtsne(as.matrix(log2(t(lengthScaledTPM.sig.genes)+1)),perplexity = p, check_duplicates = FALSE, theta=0.0 )

  tsne_plot <- data.frame(tsne_out$Y)
  tsne_plot$sample_accession <- colnames(lengthScaledTPM.sig.genes)
  
  p2 = tsne_plot %>% left_join(.,core_info)  %>%
    ggplot(.,aes(x=X1,y=X2,colour=Tissue,shape=Tissue)) + 
    geom_point(size=4) + scale_shape_manual(values=c(0:20,35:50)) +
    ggtitle(paste0("t-sne. Perplexity = ", p)) + 
    coord_cartesian(xlim = c(-50, 50), ylim = c(-50, 50))
  print(p2)
}

############################
# Retina+RPE
############################
for(p in perplexities){
  # All genes
  set.seed(47)
  tsne_out <- Rtsne(as.matrix(log2(t(lengthScaledTPM)+1)),perplexity = p, check_duplicates = FALSE, theta=0.0 )

  tsne_plot <- data.frame(tsne_out$Y)
  tsne_plot$sample_accession <- colnames(lengthScaledTPM)

  p1 = tsne_plot %>% left_join(.,core_info)  %>%
    ggplot(.,aes(x=X1,y=X2,colour=Tissue,shape=Tissue)) + 
    geom_point(size=4) + scale_shape_manual(values=c(0:20,35:50)) +
    ggtitle(paste0("t-sne. Perplexity = ", p)) + 
    coord_cartesian(xlim = c(-50, 50), ylim = c(-50, 50))
  print(p1)
  
  # RF genes removed
  sig.genes.indices = which(rownames(lengthScaledTPM) %in% sig.genes.list.retina.rpe)
  lengthScaledTPM.sig.genes = lengthScaledTPM[-(sig.genes.indices), ]
  
  set.seed(47)
  tsne_out <- Rtsne(as.matrix(log2(t(lengthScaledTPM.sig.genes)+1)),perplexity = p, check_duplicates = FALSE, theta=0.0 )

  tsne_plot <- data.frame(tsne_out$Y)
  tsne_plot$sample_accession <- colnames(lengthScaledTPM.sig.genes)
  
  p2 = tsne_plot %>% left_join(.,core_info)  %>%
    ggplot(.,aes(x=X1,y=X2,colour=Tissue,shape=Tissue)) + 
    geom_point(size=4) + scale_shape_manual(values=c(0:20,35:50)) +
    ggtitle(paste0("t-sne. Perplexity = ", p)) + 
    coord_cartesian(xlim = c(-50, 50), ylim = c(-50, 50))
  print(p2)
}
```