---
title: "WGCNA"
output:
  html_document: default
  html_notebook: default
---

# Code adapted from WGCNA tutorials 
(https://labs.genetics.ucla.edu/horvath/CoexpressionNetwork/Rpackages/WGCNA/Tutorials/index.html)

We begin by loading in the necessary packages and data. 

```{r, message = FALSE}
library(WGCNA)
library(dplyr)
library(tidyr)
library(DESeq2)
library(Rtsne)
library(ggplot2)
library(topGO)
library(sva)
library(biomaRt)

load("~/Documents/git/unified_gene_expression/data/counts_processed.Rdata")
load("~/Documents/git/unified_gene_expression/data/core_metaData.Rdata")

expr = t(counts)
meta = core_tight

dim(expr)
dim(meta)
```

We left_join the meta sheet (for the *sample_accession* variable) and the expression matrix. We then subset the resulting matrix into *RPE*, *Retina*, and *RPE+Retina* tables.

```{r, message = FALSE}
sample_accession = rownames(expr)
expr.temp = as.data.frame(cbind(sample_accession, expr))

rownames(expr.temp)[1:5]
colnames(expr.temp)[1:5]

expr.meta = left_join(expr.temp, meta, by = "sample_accession") %>% dplyr::select(-run_accession) %>% distinct()

expr.temp$sample_accession = as.character(expr.temp$sample_accession)

expr.temp$sample_accession[1:5]
expr.meta$sample_accession[1:5]
identical(expr.temp$sample_accession, expr.meta$sample_accession)

rownames(expr.meta) = expr.meta$sample_accession

expr.meta[1:5, 1:3]

class(expr.meta)
table(expr.meta$Tissue)

expr.retina = filter(expr.meta, Tissue == "Retina")
expr.rpe = filter(expr.meta, Tissue == "RPE")
expr.retina.rpe = filter(expr.meta, Tissue == "Retina" | Tissue == "RPE")

rownames(expr.retina) = expr.retina$sample_accession
rownames(expr.rpe) = expr.rpe$sample_accession
rownames(expr.retina.rpe) = expr.retina.rpe$sample_accession

expr.retina = expr.retina[, c(2:ncol(expr.retina), 1)]
expr.rpe = expr.rpe[, c(2:ncol(expr.rpe), 1)]
expr.retina.rpe = expr.retina.rpe[, c(2:ncol(expr.retina.rpe), 1)]

expr.retina[1:5, 1:5]
expr.rpe[1:5, 1:5]
expr.retina.rpe[1:5, 1:5]

dim(expr.retina)
dim(expr.rpe)
dim(expr.retina.rpe)

save(expr.retina, file = "~/Documents/git/unified_gene_expression/data/counts_retina.Rdata")
save(expr.rpe, file = "~/Documents/git/unified_gene_expression/data/counts_rpe.Rdata")
save(expr.retina.rpe, file = "~/Documents/git/unified_gene_expression/data/counts_retina_rpe.Rdata")
```

We first build the network with just the samples from the retina.

```{r, message = FALSE}
#####################################
##         Retina Network          ##
#####################################

meta.retina = expr.retina[, which(colnames(expr.retina) == "study_accession"):ncol(expr.retina)]
expr.retina = expr.retina[, 1:(which(colnames(expr.retina) == "study_accession") - 1)]

expr.retina[] = sapply(expr.retina, function(f) as.numeric(levels(f))[f])

save(expr.retina, file = "~/Documents/git/unified_gene_expression/data/expr_retina.Rdata")
save(meta.retina, file = "~/Documents/git/unified_gene_expression/data/meta_retina.Rdata")
```

We engage in some pre-processing of the data. We first filter out any genes for which at least 10% of samples have less than 10 counts. We then apply a variance stabilizing transformation and check to see that all the genes are satisfactory (in terms of variance and low expression).

```{r, message = FALSE}
min.expr.indices = which(colSums(expr.retina >= 10) >= 0.9*nrow(expr.retina))
expr.retina = expr.retina[, min.expr.indices]

expr.retina.var.stab = varianceStabilizingTransformation(round(as.matrix(expr.retina)))
save(expr.retina.var.stab, file = "~/Documents/git/unified_gene_expression/data/counts_retina_varStab.Rdata")

gsg = goodSamplesGenes(expr.retina.var.stab)
gsg$allOK
```

We perform hierarchical cluster (using average linkage) on the samples to look for any outliers. We notice that the samples are clustering based on which study they came from. 

```{r, message = FALSE}
sampleTree = hclust(dist(expr.retina.var.stab), method = "average")

# Convert traits to a color representation: white means low, red means high, grey means missing entry
traitColors = numbers2colors(as.numeric(factor(meta.retina$study_accession)), signed = FALSE);
# Plot the sample dendrogram and the colors underneath.
plotDendroAndColors(sampleTree, traitColors,
                    groupLabels = "study_accession",
                    main = "Sample dendrogram and study source (study_accesssion)")
```

We find in the dendrogram plot that there are batch effects with the different data sources. We further investigate these batch issues using t-SNE (t-Stochastic Neighbor Embedding).

```{r, message = FALSE}

set.seed(47)

perplexities = seq(10, 30, by = 10)

for(p in perplexities){
  
  tsne_out = Rtsne(as.matrix(expr.retina.var.stab), perplexity = p, check_duplicates = F, theta = 0.0)
  tsne_plot = data.frame(tsne_out$Y)
  
  p1 = tsne_plot %>%
    ggplot(aes(x=X1,y=X2,colour=factor(meta.retina$study_accession))) + 
    geom_point(size=4) + ggtitle(paste0("t-sne. Perplexity = ", p))
  print(p1)
  
  p2 = tsne_plot %>%
    ggplot(aes(x=X1,y=X2,colour=factor(meta.retina$Sub_Tissue))) + 
    geom_point(size=4) + ggtitle(paste0("t-sne. Perplexity = ", p))
  print(p2)
}
```

We then use Combat (from the **sva** package) to adjust for the batch issues (fetal vs. adult, study source). 

```{r, message = FALSE}
expr.retina.combat = as.data.frame(t(expr.retina.var.stab))

# Remove the data from studies that have very few observations
small.study.indices = which(meta.retina$study_accession %in% c("SRP002881", "SRP015336"))
expr.retina.combat = expr.retina.combat[, -(small.study.indices)]
meta.retina.combat = meta.retina[-(small.study.indices), ]

meta.retina.combat.mergeVar = meta.retina.combat %>%
  unite(study_subTissue, study_accession, Sub_Tissue)

modCombat = model.matrix(~1, data = meta.retina.combat.mergeVar)
combat_expr = t(ComBat(dat = expr.retina.combat, batch = factor(meta.retina.combat.mergeVar$study_subTissue), 
                     mod = modCombat, par.prior=TRUE, prior.plots=FALSE))
```

Following batch correction with Combat, we see that the batch issues are resolved, as clustering is minimal in t-SNE.

```{r, message = FALSE}
set.seed(47)

perplexities = seq(10, 30, by = 10)

for(p in perplexities){
  
  tsne_out = Rtsne(as.matrix(combat_expr), perplexity = p, check_duplicates = F, theta = 0.0)
  tsne_plot = data.frame(tsne_out$Y)
  
  p1 = tsne_plot %>%
    ggplot(aes(x=X1,y=X2,colour=factor(meta.retina.combat.mergeVar$study_subTissue))) + 
    geom_point(size=4) + ggtitle(paste0("t-sne (post-Combat adjustment). Perplexity = ", p))
  print(p1)
  
  p2 = tsne_plot %>%
    ggplot(aes(x=X1,y=X2,colour=factor(meta.retina.combat.mergeVar$study_subTissue))) + 
    geom_point(size=4) + ggtitle(paste0("t-sne (post-Combat adjustment). Perplexity = ", p))
  print(p2)
}
```

Following batch correction, we see that the distinct clusters are no longer present. Now we recluster the samples with hierarchical clustering (average linkage). The dendrogram shows that the batch issues have been resolved (at least partially). 

```{r, message = FALSE}
sampleTree.combat = hclust(dist(combat_expr), method = "average")

# Convert traits to a color representation: white means low, red means high, grey means missing entry
traitColors = numbers2colors(as.numeric(factor(meta.retina.combat.mergeVar$study_subTissue)), signed = FALSE);
# Plot the sample dendrogram and the colors underneath.
plotDendroAndColors(sampleTree.combat, traitColors,
                    groupLabels = "study_accession",
                    main = "Sample dendrogram and trait heatmap")

```

Now that we have resolved the batch effect issue, we can resume building a WGCNA network, exploring both signed and unsigned networks. We first begin with automatic network building.

```{r, message = FALSE}
options(stringsAsFactors = FALSE)

####################################
##           Unsigned             ##
####################################

# Choose a set of soft-thresholding powers
powers.unsigned = c(c(1:10), seq(from = 12, to=30, by=2))
# Call the network topology analysis function
sft.unsigned = pickSoftThreshold(combat_expr, powerVector = powers.unsigned, networkType = "unsigned")

# With a power of 14 we get
# Power SFT.R.sq  slope truncated.R.sq mean.k. median.k. max.k.
# 14     0.8070   -1.530        0.950   70.90    31.600  502.0

####################################
##           Signed             ##
####################################

# Choose a set of soft-thresholding powers
powers.signed = c(c(1:10), seq(from = 12, to=30, by=2))
# Call the network topology analysis function
sft.signed = pickSoftThreshold(combat_expr, powerVector = powers.signed, networkType = "signed")

# With a power of 30 we get
# Power SFT.R.sq  slope truncated.R.sq mean.k. median.k. max.k.
# 30    0.8030    -1.550      0.950    70.7      34.0    478

```

We achieve scale-free topology indices of 0.8 for powers of 14 and 30 for unsigned and signed networks, respectively. We build the networks using these power parameters. Upon examining the resulting gene dendrograms, it's clear that the unsigned network is superior.

```{r, message = FALSE}
net.unsigned = blockwiseModules(combat_expr, power = 14, maxBlockSize = 14000,
                       TOMType = "unsigned", minModuleSize = 30,
                       reassignThreshold = 0, mergeCutHeight = 0.1,
                       numericLabels = TRUE, pamRespectsDendro = FALSE,
                       verbose = 3)

net.signed = blockwiseModules(combat_expr, power = 30, maxBlockSize = 14000,
                       TOMType = "signed", minModuleSize = 30,
                       reassignThreshold = 0, mergeCutHeight = 0.05,
                       numericLabels = TRUE, pamRespectsDendro = FALSE,
                       verbose = 3)

save(net.unsigned, file = "~/Documents/git/unified_gene_expression/data/retina_net_unsigned.Rdata")
save(net.signed, file = "~/Documents/git/unified_gene_expression/data/retina_net_signed.Rdata")

# Examining number of modules
table(net.unsigned$colors)
# Convert labels to colors for plotting
mergedColors = labels2colors(net.unsigned$colors)
# Plot the dendrogram and the module colors underneath
plotDendroAndColors(net.unsigned$dendrograms[[1]], mergedColors[net.unsigned$blockGenes[[1]]],
                    "Module colors",
                    dendroLabels = FALSE, hang = 0.03,
                    addGuide = TRUE, guideHang = 0.05)

# Examining number of modules
table(net.signed$colors)
# Convert labels to colors for plotting
mergedColors = labels2colors(net.signed$colors)
# Plot the dendrogram and the module colors underneath
plotDendroAndColors(net.signed$dendrograms[[1]], mergedColors[net.signed$blockGenes[[1]]],
                    "Module colors",
                    dendroLabels = FALSE, hang = 0.03,
                    addGuide = TRUE, guideHang = 0.05)

```

# Building a network step-by-step

We now consider building a network using the TOM (Topological Overlap Matrix) and hierarchical clustering, as opposed to the automatic network construction performed above.

```{r, message = FALSE}
softPower = 14
adjacency = adjacency(combat_expr, power = softPower)

# Turn adjacency into topological overlap
TOM = TOMsimilarity(adjacency)
dissTOM = 1-TOM

# Call the hierarchical clustering function
geneTree = hclust(as.dist(dissTOM), method = "average");
# Plot the resulting clustering tree (dendrogram)
plot(geneTree, xlab="", sub="", main = "Gene clustering on TOM-based dissimilarity",
     labels = FALSE, hang = 0.04)



# We like large modules, so we set the minimum module size relatively high:
minModuleSize = 30;
# Module identification using dynamic tree cut:
dynamicMods = cutreeDynamic(dendro = geneTree, distM = dissTOM,
                deepSplit = 2, pamRespectsDendro = FALSE,
                minClusterSize = minModuleSize);
table(dynamicMods)

# Convert numeric lables into colors
dynamicColors = labels2colors(dynamicMods)
table(dynamicColors)
# Plot the dendrogram and colors underneath
plotDendroAndColors(geneTree, dynamicColors, "Dynamic Tree Cut",
                    dendroLabels = FALSE, hang = 0.0005,
                    addGuide = TRUE, guideHang = 0.05,
                    main = "Gene dendrogram and module colors")

```

We then calculate the eigengenes for the different modules.

```{r, message = FALSE}
# Calculate eigengenes
MEList = moduleEigengenes(combat_expr, colors = dynamicColors)
MEs = MEList$eigengenes
# Calculate dissimilarity of module eigengenes
MEDiss = 1-cor(MEs);
# Cluster module eigengenes
METree = hclust(as.dist(MEDiss), method = "average");
# Plot the result
plot(METree, main = "Clustering of module eigengenes",
     xlab = "", sub = "")

MEDissThres = 0.05   # used to be 0.15
# Plot the cut line into the dendrogram
abline(h=MEDissThres, col = "red")
# Call an automatic merging function
merge = mergeCloseModules(combat_expr, dynamicColors, cutHeight = MEDissThres, verbose = 3)
# The merged module colors
mergedColors = merge$colors
# Eigengenes of the new merged modules:
mergedMEs = merge$newMEs

```


```{r, message = FALSE}
plotDendroAndColors(geneTree, cbind(dynamicColors, mergedColors),
                    c("Dynamic Tree Cut", "Merged dynamic"),
                    dendroLabels = FALSE, hang = 0.03,
                    addGuide = TRUE, guideHang = 0.05)
```

```{r, message = FALSE}
# Rename to moduleColors
moduleColors = mergedColors
# Construct numerical labels corresponding to the colors
colorOrder = c("grey", standardColors(50));
moduleLabels = match(moduleColors, colorOrder)-1;
MEs = mergedMEs;
# Save module colors and labels for use in subsequent parts
save(MEs, moduleLabels, moduleColors, geneTree, file = "~/Documents/git/unified_gene_expression/data/networkConstruction-stepByStep.RData")
```


# Building a network step-by-step

We now build a signed network.

```{r, message = FALSE}
powers = c(1:10, seq(from = 12, to = 20, by = 2))
sft = pickSoftThreshold(combat_expr, powerVector = powers, verbose = 5)

{# Scale-free topology fit index as a function of the soft-thresholding power
plot(sft$fitIndices[,1], -sign(sft$fitIndices[,3])*sft$fitIndices[,2],
xlab="Soft Threshold (power)",ylab="Scale Free Topology Model Fit,signed R^2",type="n",
main = paste("Scale independence"));
text(sft$fitIndices[,1], -sign(sft$fitIndices[,3])*sft$fitIndices[,2], labels=powers,col="red");
# this line corresponds to using an R^2 cut-off of h
abline(h=0.80,col="red")}
# Mean connectivity as a function of the soft-thresholding power
{plot(sft$fitIndices[,1], sft$fitIndices[,5],
xlab="Soft Threshold (power)",ylab="Mean Connectivity", type="n",
main = paste("Mean connectivity"))
text(sft$fitIndices[,1], sft$fitIndices[,5], labels=powers,col="red")}

softPower = 14
adjacency = adjacency(combat_expr, power = softPower)

# Turn adjacency into topological overlap
TOM = TOMsimilarity(adjacency)
dissTOM = 1-TOM

# Call the hierarchical clustering function
geneTree = hclust(as.dist(dissTOM), method = "average");
# Plot the resulting clustering tree (dendrogram)
plot(geneTree, xlab="", sub="", main = "Gene clustering on TOM-based dissimilarity",
     labels = FALSE, hang = 0.04)


# We like large modules, so we set the minimum module size relatively high:
minModuleSize = 30;
# Module identification using dynamic tree cut:
dynamicMods = cutreeDynamic(dendro = geneTree, distM = dissTOM,
                deepSplit = 2, pamRespectsDendro = FALSE,
                minClusterSize = minModuleSize);
table(dynamicMods)

# Convert numeric lables into colors
dynamicColors = labels2colors(dynamicMods)
table(dynamicColors)
# Plot the dendrogram and colors underneath
plotDendroAndColors(geneTree, dynamicColors, "Dynamic Tree Cut",
                    dendroLabels = FALSE, hang = 0.0003,
                    addGuide = TRUE, guideHang = 0.05,
                    main = "Gene dendrogram and module colors")

```

We then calculate the eigengenes.

```{r, message = FALSE}
# Calculate eigengenes
MEList = moduleEigengenes(combat_expr, colors = dynamicColors)
MEs = MEList$eigengenes
# Calculate dissimilarity of module eigengenes
MEDiss = 1-cor(MEs);
# Cluster module eigengenes
METree = hclust(as.dist(MEDiss), method = "average");
# Plot the result
{plot(METree, main = "Clustering of module eigengenes",
     xlab = "", sub = "")

MEDissThres = 0.1   # used to be 0.15
# Plot the cut line into the dendrogram
abline(h=MEDissThres, col = "red")}
# Call an automatic merging function
merge = mergeCloseModules(combat_expr, dynamicColors, cutHeight = MEDissThres, verbose = 3)
# The merged module colors
mergedColors = merge$colors
table(mergedColors)
# Eigengenes of the new merged modules:
mergedMEs = merge$newMEs

```

```{r, message = FALSE}
plotDendroAndColors(geneTree, cbind(dynamicColors, mergedColors),
                    c("Dynamic Tree Cut", "Merged dynamic"),
                    dendroLabels = FALSE, hang = 0.03,
                    addGuide = TRUE, guideHang = 0.05)
```

# Network Visualization

We obtain the TOM and visualize the network using a heat map. Plotting on all genes is very computationally intensive, so here we just compute the TOM on a randomly selected subset of 400 genes.

```{r, message = FALSE}
# Transform dissTOM with a power to make moderately strong connections more visible in the heatmap
plotTOM = dissTOM^7;
# Set diagonal to NA for a nicer plot
diag(plotTOM) = NA;
# Call the plot function

# plotting on 
#TOMplot(plotTOM, geneTree, mergedColors, main = "Network heatmap plot, all genes")


nSelect = 400
# For reproducibility, we set the random seed
set.seed(47);
select = sample(ncol(combat_expr), size = nSelect);
selectTOM = dissTOM[select, select];
# There’s no simple way of restricting a clustering tree to a subset of genes, so we must re-cluster.
selectTree = hclust(as.dist(selectTOM), method = "average")
selectColors = mergedColors[select];
# Taking the dissimilarity to a power, say 10, makes the plot more informative by effectively changing
# the color palette; setting the diagonal to NA also improves the clarity of the plot
plotDiss = selectTOM^7;
diag(plotDiss) = NA;
TOMplot(plotDiss, selectTree, selectColors, main = "Network heatmap plot, selected genes")
```

# Pathway Analysis (Retina Network)

```{r, message = FALSE}
library(GO.db)
library(AnnotationDbi)
library(org.Hs.eg.db)


# Load annotation file
load("~/Documents/git/unified_gene_expression/data/gencode_v25_annotation.Rdata")
# Read in the probe annotation
annot = gencode_v25_annotation %>% dplyr::select(-Transcript.ID) %>% distinct()
n_distinct(annot$Gene.ID)
n_distinct(annot$Gene.Name)

# Removing alternative splicing artifacts
annot[, 'Gene.ID'] = gsub("\\.[^.]*$", "", annot$Gene.ID)
annot = annot %>% distinct()
dim(annot)

# Match probes in the data set to the probe IDs in the annotation file 
probes = colnames(combat_expr)
probes2annot = which(annot$Gene.Name %in% probes)
# Get the corresponding Gene IDs
allGeneIDs = annot[probes2annot, ]
# (For now - need new method) eliminate redundancy for gene names
allGeneIDs = allGeneIDs[!duplicated(allGeneIDs$Gene.Name),]
dim(allGeneIDs)


mart <- useMart("ensembl")
datasets <- listDatasets(mart)
mart<-useDataset("hsapiens_gene_ensembl",mart)
entrez_ids = getBM(attributes=c("ensembl_gene_id", "entrezgene"),filters="ensembl_gene_id",values=allGeneIDs, mart=mart)

test = merge(allGeneIDs,entrez_ids,by.x="Gene.ID",by.y="ensembl_gene_id") %>% distinct()

moduleColors = labels2colors(net.unsigned$colors)

# $ Choose interesting modules
intModules = c("brown", "green", "red")
for (module in intModules)
{
  # Select module probes
  modGenes = (moduleColors==module)
  # Get their entrez ID codes
  modGeneIDs = allGeneIDs[modGenes];
  # Write them into a file
  fileName = paste("~/Documents/git/unified_gene_expression/results/GeneIDs-", module, ".txt", sep="");
  write.table(as.data.frame(modGeneIDs), file = fileName,
              row.names = FALSE, col.names = FALSE)
}
# As background in the enrichment analysis, we will use all probes in the analysis.
fileName = paste("~/Documents/git/unified_gene_expression/results/GeneIDs-all.txt", sep="");
write.table(as.data.frame(allGeneIDs), file = fileName,
            row.names = FALSE, col.names = FALSE)
```

**Issue to resolve:** converting from ensembl gene IDs to entrez gene IDs.

```{r, message = FALSE}
#GOenr = GOenrichmentAnalysis(moduleColors, allGeneIDs, organism = "human", nBestP = 10)
```

# Moving forward
* Build network on RPE tissue samples.
* Consider whether it's appropriate to correct for fetal vs. adult.
* Identify the goal for pathway analysis.
* Compare advantages/disadvantages of different networks considered.
* Compare Retina and RPE networks.
    + PPT: https://labs.genetics.ucla.edu/horvath/CoexpressionNetwork/WORKSHOP/2013/Miller/Miller1_ComparingMouseAndHumanBrain.pdf
    + Lecture: https://www.youtube.com/watch?v=aBD67YmCBK4


